---
title: "Patient-level spatial mIF workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Patient-level spatial mIF workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(spatialMIFtools)
set.seed(20260803)
phenotypes <- c("Epithelial", "APC", "Treg", "CTL", "HelperT")
```

## Sampling unit and input contract

One row must represent one physical cell. `patient` is the independent sampling
unit; `image` identifies repeated fields; `group` defines the pathology
contrast. Coordinates are numeric and phenotype calls are binary. Overlapping
positive calls are allowed when they are part of the supplied annotation.

```{r data}
cells <- simulate_mif_cells(12, 2, 180)
validate_mif_cells(cells, phenotypes)
```

## Grids and epithelial-density zones

```{r grid}
grids <- bin_mif_cells(cells, phenotypes, nx = 12, ny = 9, min_cells = 5)
zones <- add_epithelial_zones(grids, epithelial = "Epithelial", sigma = 1.25)
table(zones$tumor_zone)
```

The density variables are log counts per fixed image area, not within-bin
proportions. Q1--Q4 are computed from within-image ranks of a smoothed
epithelial count surface. They are not absolute thresholds or temporal stages.

## Neighborhoods and conditional networks

```{r neighborhoods}
neighborhoods <- fit_spatial_neighborhoods(zones, phenotypes, k_range = 3:5)
assigned <- assign_neighborhoods(neighborhoods, zones)
patient_neighborhood_prevalence(assigned)
```

```{r networks}
networks <- estimate_ridge_networks(assigned, phenotypes)
adjusted <- estimate_ridge_networks(
  assigned, phenotypes, adjust_cellularity = TRUE
)
patient_edge_contrasts(adjusted, c("Group1", "Group2"))
```

An adjusted edge conditions on the other measured phenotype-density fields and
on the specified linear total-cellularity nuisance model. It remains a
statistical association.

## Fully nested prediction

```{r prediction, eval=FALSE}
scores <- nested_lopo_prediction(
  assigned,
  phenotypes,
  group_levels = c("Group1", "Group2"),
  feature_set = "combined",
  k_range = 3:5
)
attr(scores, "auc")
```

Every LOPO fold relearns density scaling, K, centers, held-out assignments,
predictor scaling, and ridge-logistic coefficients. The result is internal
patient-level discrimination, not external clinical validation.

## Optional Python backend

```{r python, eval=FALSE}
if (python_backend_available()) {
  py <- spatial_mif_python()
  py$partial_correlation(matrix(rnorm(500), ncol = 5), ridge = 0.15)
}
```
